---
title: "Data download and preprocessing"
author: "Shixiang Wang"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    lightbox: false
    toc_depth: 3
    mathjax: true
---

```{r pre-process-setup, include=FALSE}
options(max.print="75")
knitr::opts_chunk$set(echo = TRUE, comment = "#>", eval = TRUE, collapse = TRUE)
knitr::opts_knit$set(width=75)
```

This part will clearly describe how to process raw data but not provide an easy script for re-running all preprocess steps just by a click due to the complexity of data preprocessing.

## Dependencies

The preprocessing step is dependent on R software, if you have not used R yet or R is not installed, please see <https://cran.r-project.org/>.

R packages:

* [UCSCXenaTools](https://github.com/ShixiangWang/UCSCXenaTools)
* [tidyverse](https://www.tidyverse.org/)
* [TCGAmutations](https://github.com/PoisonAlien/TCGAmutations)
* [DT](https://cran.r-project.org/web/packages/DT/index.html)
* [GSVA](https://github.com/rcastelo/GSVA)

## TCGA Pan-cancer data download

TCGA Pan-cancer data (Version 2017-10-13), including datasets of clinical informaiton, gene expression, are downloaded from [UCSC Xena](https://xenabrowser.net/datapages/) via R package [UCSCXenaTools](https://github.com/ShixiangWang/UCSCXenaTools).

If you have not installed this package yet, please run following command in R.

```{r install_ucscxenatools, eval=FALSE}
install.packages("UCSCXenaTools")
```

Load the package with:

```{r load_xenatools, message=FALSE}
library(UCSCXenaTools)

# also load dplyr package
library(dplyr)
```


Obtain datasets information available at [TCGA data hubs of UCSC Xena](https://xenabrowser.net/datapages/?host=https%3A%2F%2Ftcga.xenahubs.net&removeHub=https%3A%2F%2Fxena.treehouse.gi.ucsc.edu%3A443):

```{r}
xe = XenaGenerate(subset = XenaHostNames == "TCGA")
xe
```


Obtain clinical datasets of TCGA:

```{r query_pan_cli}
xe %>% XenaFilter(filterDatasets = "clinical") -> xe_clinical
xe_clinical
```

Obtain gene expression datasets of TCGA:

```{r query_pan_ge}
xe %>% XenaFilter(filterDatasets = "HiSeqV2_PANCAN$") -> xe_rna_pancan
xe_rna_pancan
```

Create data queries and download them:

```{r download_xena_pancan, eval=FALSE}
xe_clinical.query = XenaQuery(xe_clinical)
xe_clinical.download = XenaDownload(xe_clinical.query, 
                                    destdir = "UCSC_Xena/TCGA/Clinical")

xe_rna_pancan.query = XenaQuery(xe_rna_pancan)
xe_rna_pancan.download = XenaDownload(xe_rna_pancan.query, 
                                      destdir = "UCSC_Xena/TCGA/RNAseq_Pancan")
```

```{r hide_download_pancan, include=FALSE}
if (!dir.exists("UCSC_Xena")){
    xe_clinical.query = XenaQuery(xe_clinical)
    xe_clinical.download = XenaDownload(xe_clinical.query, 
                                        destdir = "UCSC_Xena/TCGA/Clinical")
    
    xe_rna_pancan.query = XenaQuery(xe_rna_pancan)
    xe_rna_pancan.download = XenaDownload(xe_rna_pancan.query, 
                                          destdir = "UCSC_Xena/TCGA/RNAseq_Pancan")
}
```


The RNASeq data we downloaded are pancan normalized. 

> For comparing data within independent cohort (like TCGA-LUAD), we recommend to use the "gene expression RNAseq" dataset. For questions regarding the gene expression of this particular cohort in relation to other types tumors, you can use the pancan normalized version of the "gene expression RNAseq" data. For comparing with data outside TCGA, we recommend using the percentile version if the non-TCGA data is normalized by percentile ranking. For more information, please see our Data FAQ: [here](https://docs.google.com/document/d/1q-7Tkzd7pci4Rz-_IswASRMRzYrbgx1FTTfAWOyHbmk/edit?usp=sharing).

These datasets are downloaded to local machine, we need to load them into R. However, whether filenames of datasets or contents in datasets all look messy, next we need to clean them before real analysis.

## TCGA pan-cancer data clean

### Clean filenames

First, clean filenames.

```{r clean_filenames}
# set data directory where TCGA clinical and pancan RNAseq data stored
TCGA_DIR = "UCSC_Xena/TCGA" 
# obtain filenames of rna-seq data
dir(paste0(TCGA_DIR,"/RNAseq_Pancan")) -> RNAseq_filelist
# obtain tcga project code
sub("TCGA\\.(.*)\\.sampleMap.*", "\\1", RNAseq_filelist) -> project_code
# obtain filenames of clinical datasets
dir(paste0(TCGA_DIR,"/Clinical")) -> Clinical_filelist
```

Check.

```{r}
head(RNAseq_filelist)

head(project_code)

head(Clinical_filelist)
```

Obtain TCGA project abbreviations from <https://gdc.cancer.gov/resources-tcga-users/tcga-code-tables/tcga-study-abbreviations> and read as a `data.frame` in R.

```{r}
library(readr)
TCGA_Study = read_tsv("
                      LAML	Acute Myeloid Leukemia
                      ACC	Adrenocortical carcinoma
                      BLCA	Bladder Urothelial Carcinoma
                      LGG	Brain Lower Grade Glioma
                      BRCA	Breast invasive carcinoma 
                      CESC	Cervical squamous cell carcinoma and endocervical adenocarcinoma
                      CHOL	Cholangiocarcinoma
                      LCML	Chronic Myelogenous Leukemia
                      COAD	Colon adenocarcinoma
                      CNTL	Controls
                      ESCA	Esophageal carcinoma
                      FPPP	FFPE Pilot Phase II
                      GBM	Glioblastoma multiforme
                      HNSC	Head and Neck squamous cell carcinoma
                      KICH	Kidney Chromophobe
                      KIRC	Kidney renal clear cell carcinoma
                      KIRP	Kidney renal papillary cell carcinoma
                      LIHC	Liver hepatocellular carcinoma
                      LUAD	Lung adenocarcinoma
                      LUSC	Lung squamous cell carcinoma
                      DLBC	Lymphoid Neoplasm Diffuse Large B-cell Lymphoma
                      MESO	Mesothelioma
                      MISC	Miscellaneous
                      OV	Ovarian serous cystadenocarcinoma
                      PAAD	Pancreatic adenocarcinoma
                      PCPG	Pheochromocytoma and Paraganglioma
                      PRAD	Prostate adenocarcinoma
                      READ	Rectum adenocarcinoma
                      SARC	Sarcoma
                      SKCM	Skin Cutaneous Melanoma
                      STAD	Stomach adenocarcinoma
                      TGCT	Testicular Germ Cell Tumors
                      THYM	Thymoma
                      THCA	Thyroid carcinoma
                      UCS	Uterine Carcinosarcoma
                      UCEC	Uterine Corpus Endometrial Carcinoma
                      UVM	Uveal Melanoma")
colnames(TCGA_Study) = c("StudyAbbreviation", "StudyName")
```

Compare difference.

```{r}
intersect(project_code, TCGA_Study$StudyAbbreviation) -> project_code
Clinical_filelist[rowSums(sapply(paste0("__",project_code,"_"), function(x) grepl(x, Clinical_filelist)))>0] -> Clinical_filelist2

setdiff(Clinical_filelist, Clinical_filelist2)
```

Remove `TCGA.FPPP.sampleMap__FPPP_clinicalMatrix.gz` which has no RNAseq data. Remove other 3 datasets which merge more than one TCGA study, only keep individual study from TCGA.


### Clean clinical datasets

Now, we read TCGA clinical datasets and clean them.

```{r clean_tcga_cli}
library(tidyverse)
# keep valuable columns of clinical datasets
select_cols = c("sampleID", "OS", "OS.time", "OS.unit", "RFS", "RFS.time", "RFS.unit", 
                "age_at_initial_pathologic_diagnosis", "gender", "tobacco_smoking_history",
                "tobacco_smoking_history_indicator", "sample_type", "pathologic_M",
                "pathologic_N", "pathologic_T", "pathologic_stage")

# read clinical files
Cli_DIR = paste0(TCGA_DIR,"/Clinical")

#----------------------------------
# Load and preprocess clinical data
#----------------------------------
Clinical_List = XenaPrepare(paste0(Cli_DIR,"/",Clinical_filelist2))
names(Clinical_List)
sub("TCGA\\.(.*)\\.sampleMap.*", "\\1", names(Clinical_List)) -> project_code

TCGA_Clinical = tibble()
# for loop
for (i in 1:length(project_code)){
    clinical = names(Clinical_List)[i]
    project = project_code[i]
    df = Clinical_List[[clinical]]
    col_exist = select_cols %in% colnames(df)
    res = tibble()
    if(!all(col_exist)){
        res = df[, select_cols[col_exist]]
        res[, select_cols[!col_exist]] = NA
    }else{
        res = df[, select_cols]
    }
    res$Project = project
    res %>% select(Project, select_cols) -> res
    TCGA_Clinical = bind_rows(TCGA_Clinical, res)
}

rm(res, df, i, clinical, project, col_exist);gc() # remove temp variables
```

View the clinical data, check variables.

```{r view_pan_cli, warning=FALSE}
DT::datatable(TCGA_Clinical, 
              options = list(scrollX = TRUE, keys = TRUE), rownames = FALSE)
```

All unit are same in days, remove two column with redundant information.

```{r}
TCGA_Clinical = TCGA_Clinical %>% select(-c(OS.unit, RFS.unit))
```


Continue to tidy clinical datasets: rename variables, filter 14 samples with unusual sample types, rename variable value and . After cleanning, save the result object `TCGA_Clinical.tidy` for following analysis.

```{r, eval=FALSE}
# create a new tidy dataframe
TCGA_Clinical.tidy = TCGA_Clinical %>%
    rename(Age = age_at_initial_pathologic_diagnosis, Gender = gender, 
                         Smoking_history = tobacco_smoking_history,
                         Smoking_indicator = tobacco_smoking_history_indicator,
                         Tumor_Sample_Barcode = sampleID) %>% 
    filter(sample_type %in% c("Solid Tissue Normal", "Primary Tumor", "Metastatic", "Recurrent Tumor",
                              "Primary Blood Derived Cancer - Peripheral Blood")) %>% # Additional - New Primary, Additional Metastatic, FFPE Scrolls total 14 sample removed
    mutate(Gender = case_when(
        Gender == "FEMALE" ~ "Female",
        Gender == "MALE"   ~ "Male",
        TRUE               ~ NA_character_
    ), Tumor_stage = case_when(
        pathologic_stage == "Stage 0" ~ "0",
        pathologic_stage %in% c("Stage I", "Stage IA", "Stage IB") ~ "I",
        pathologic_stage %in% c("Stage II", "Stage IIA", "Stage IIB", "Stage IIC") ~ "II",
        pathologic_stage %in% c("Stage IIIA", "Stage IIIB", "Stage IIIC") ~ "III",
        pathologic_stage %in% c("Stage IV", "Stage IVA", "Stage IVB", "Stage IVC") ~ "IV",
        pathologic_stage == "Stage X" ~ "X",
        TRUE ~ NA_character_)) %>% 
    mutate(Gender = factor(Gender, levels = c("Male", "Female")), 
           Tumor_stage = factor(Tumor_stage, levels = c("0", "I", "II", "III", "IV", "X"))) 

if(!file.exists("results/TCGA_tidy_Clinical.RData")){
    dir.create("results", showWarnings = FALSE)
    save(TCGA_Clinical.tidy, file = "results/TCGA_tidy_Clinical.RData")
}   
```

### Clean RNASeq datasets

Now, we read TCGA pan-cancer RNASeq datasets and clean them.

```{r clean_pan_rnaseq, eval=FALSE}
dir(paste0(TCGA_DIR,"/RNAseq_Pancan")) -> RNAseq_filelist.Pancan
RNAseq_filelist.Pancan[rowSums(sapply(paste0("TCGA.",project_code,".sampleMap"), function(x) grepl(x, RNAseq_filelist.Pancan)))>0] -> RNAseq_filelist.Pancan2

RNASeqDIR.pancan = paste0(TCGA_DIR,"/RNAseq_Pancan")
RNASeq_List.pancan = XenaPrepare(paste0(RNASeqDIR.pancan,"/",RNAseq_filelist.Pancan2))
names(RNASeq_List.pancan)
sapply(RNASeq_List.pancan, function(x)nrow(x))

names(RNASeq_List.pancan) = sub("TCGA\\.(.*)\\.sampleMap.*", "\\1", names(RNASeq_List.pancan))

RNASeq_pancan = purrr::reduce(RNASeq_List.pancan, full_join)
if(!file.exists("results/TCGA_RNASeq_PanCancer.RData")){
    save(RNASeq_pancan, file = "results/TCGA_RNASeq_PanCancer.RData")
}


#class(RNASeq_pancan)
rm(RNASeq_List.pancan);gc()
```

The result object `RNASeq_pancan` is very huge, it needs about 1.5GB space to store. We will not view it as a table in html page.

## Selection of APM genes and marker genes of immune cell type

Marker genes for immune cell types were obtained from [Senbabaoglu, Y. et al](https://www.ncbi.nlm.nih.gov/pubmed/27855702), selection of genes involved in processing and presentation of antigen (APM) was also inspired by this paper and finally 18 genes were selected based on [literature review of APM](https://academic.oup.com/jnci/article-lookup/doi/10.1093/jnci/djt184) and some facts that mutations or deletions affecting genes encoding APM (HLA, beta2-microglobulin, TAP1/2 etc.) may lead to __reduced presentation of neoantigens__ by a cancer cell.

```{r immune_genes, message=FALSE}
APM_genes = read_csv("../data/APM.csv", skip = 1)

immune_cellType = read_csv("../data/Immune_Cell_type_List.csv", skip=1)
immune_cellType = immune_cellType %>% filter(inRNAseq == "YES")

# merge two gene list
merged_geneList = bind_rows(immune_cellType,
          tibble(Cell_type="APM", Symbol=APM_genes$Gene_Name,
                                 Name = APM_genes$Protein_Name, inRNAseq = "YES")) 

# save data
save(merged_geneList, file = "results/merged_geneList.RData")
```

View gene list.

```{r view_merged_genelist}
DT::datatable(merged_geneList, 
              options = list(scrollX = TRUE, keys = TRUE), rownames = FALSE)
```


Remove all variables.

```{r}
rm(list=ls());gc()
```



## Calculation of APS, TIS and IIS

> The GSVA package allows one to perform a change in coordinate systems of molecular measurements, transforming the data from a gene by sample matrix to a gene-set by sample matrix, thereby allowing the evaluation of pathway enrichment for each sample. This new matrix of GSVA enrichment scores facilitates applying standard analytical methods like functional enrichment, survival analysis, clustering, CNV-pathway analysis or cross-tissue pathway analysis, in a pathway-centric manner. 

APM score (APS) and score of each immune cell type are directly calculated from GSVA method. Aggregate TIS (T cell infiltration score) and IIS (Immune cell infiltration score) are calculated using following method, their effectiveness have been validated by many studies. 

> The TIS was defined as the mean of the standardized values for the following T cell subsets: CD8 T, T helper, T, T central and effector memory, Th1, Th2, Th17, and Treg cells. The immune infiltration score (IIS) for a sample was defined as the mean of standardized values for macrophages, DC subsets (total, plasmacytoid, immature, activated), B cells, cytotoxic cells, eosinophils, mast cells, neutrophils, NK cell subsets (total, CD56 bright, CD56 dim), and all T cell subsets (CD8 T, T helper, T central and effector memory, Th1, Th2, Th17, and Treg cells). ([Senbabaoglu, Y. et al](https://www.ncbi.nlm.nih.gov/pubmed/27855702))

Load function used to apply GSVA method.

```{r}
source("../code/functions.R")
```

Apply GSVA to TCGA data. __This step should run on a machine with big memory.__

```{r, eval=FALSE} 
# load data
load("results/merged_geneList.RData")
load("results/TCGA_RNASeq_PanCancer.RData")


applyGSVA(merged_geneList, group_col = "Cell_type", gene_col = "Symbol", ExprMatList = list(RNASeq_pancan), method = "gsva") -> res_pancan.GSVA

save(res_pancan.GSVA, file="results/res_pancan.GSVA.RData")
```

Calculate TIS and IIS.

```{r calc_TIS_IIS}
if (!exists("res_pancan.GSVA")) {
    load("results/res_pancan.GSVA.RData")
}

res_pancan.GSVA = res_pancan.GSVA[[1]]

calc_TisIIs = function(df){
    df %>% 
        mutate(TIS = (`CD8 T cells` + `T helper cells` + `T cells` + `Tcm cells` + `Tem cells` + `Th1 cells` + `Th2 cells` + `Th17 cells` + `Treg cells`) / 9,
               IIS = (`CD8 T cells` + `T helper cells` + `T cells` + `Tcm cells` + `Tem cells` + `Th1 cells` + `Th2 cells` + `Th17 cells` + `Treg cells` + aDC + `B cells` + `Cytotoxic cells` + DC + Eosinophils + iDC + Macrophages + `Mast cells` + Neutrophils + `NK CD56bright cells` + `NK CD56dim cells` + `NK cells` + pDC) / 22) -> df
}

# tidy res_pancan.GSVA
res_pancan.GSVA = rownames_to_column(res_pancan.GSVA, var = "tsb")

# calculate TIS and IIS
gsva.pac = calc_TisIIs(res_pancan.GSVA)
```


View this table (head 100 rows).

```{r view_GSVA}
DT::datatable(head(gsva.pac, 100), 
              options = list(scrollX = TRUE, keys = TRUE), rownames = FALSE)
```


save GSVA scores of APM and immune cell types, TIS and IIS.

```{r}
if (!file.exists("results/gsva_tcga_pancan.RData")){
    save(gsva.pac, file ="results/gsva_tcga_pancan.RData")
}

rm(list = ls()); gc()
```


## TIMER data download and clean

TIMER is a method that can accurately resolve relative fractions of diverse cell types based on gene expression profiles from complex tissues. We downloaded TIMER result for TCGA samples from [Li, B. et al.](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-016-1028-7).

Load TIMER result and clean it for use.

```{r clean_timer_result, message=FALSE}
timer = read_tsv("../data/TCGA_sixcell_population_TIMER.txt")
timer$sample %>% length()
timer$sample %>% unique() %>% length()
timer$sample %>% substr(start=1, stop=12) %>% unique() %>% length()

# keep only tumor samples
timer %>% filter(as.numeric(substr(sample, 14, 15)) %in% seq(1,9)) %>% 
    mutate(sample = substr(sample, 1, 15)) -> timer_clean
timer_clean %>% arrange(desc(T_cell.CD8)) %>% distinct(sample, .keep_all = TRUE) -> timer_clean

# save
if (!file.exists("results/timer.RData")){
    save(timer_clean, file = "results/timer.RData")
}

rm(list = ls()); gc()
```


## TCGA mutation download and clean

We downloaded TCGA somatic mutations (data source: [TCGA MC3](https://gdc.cancer.gov/about-data/publications/mc3-2017)) via TCGAmutations package.

> TCGAmutations is an R data package containing somatic mutations from TCGA cohorts. This is particularly useful for those working with mutation data from TCGA studies - where most of the time is spent on searching various databases, downloading, compiling and tidying up the data before even the actual analysis is started. This package tries to mitigate the issue by providing pre-compiled, curated somatic mutations from 33 TCGA cohorts along with relevant clinical information for all sequenced samples.

Load mutations of TCGA studies and merge them into one.

```{r load_tcga_mutations, eval=FALSE, warning=FALSE, message=FALSE}
## install package please use following command
# devtools::install_github(repo = "PoisonAlien/TCGAmutations")

# load package
require(TCGAmutations)


study_list = tcga_available()$Study_Abbreviation[-34]
cohorts = system.file('extdata', 'cohorts.txt', package = 'TCGAmutations')
cohorts = data.table::fread(input = cohorts)

# calculate TMB
lapply(study_list, function(study){
    require(maftools)
        TCGAmutations::tcga_load(study)
        maf = eval(as.symbol(tolower(paste0("TCGA_", study, "_mc3"))))
        maf.silent <- maf@maf.silent
        sample.silent <- maf.silent[,.N, .(Tumor_Sample_Barcode)]
        sample.nonsilent <- getSampleSummary(maf)
        res <- dplyr::full_join(sample.silent, sample.nonsilent, by="Tumor_Sample_Barcode")
        res = res %>% dplyr::mutate(TMB_Total=ifelse(!is.na(N), N+total, total), 
                              TMB_NonsynSNP=Missense_Mutation+Nonsense_Mutation,
                              TMB_NonsynVariants=total) %>% 
            dplyr::select(TMB_Total:TMB_NonsynVariants, Tumor_Sample_Barcode)   
        res
    
}) -> tcga_tmb
names(tcga_tmb) = study_list

# 33 study available, merge them
TCGA_TMB = purrr::reduce(tcga_tmb, bind_rows)

if (!file.exists("results/TCGA_TMB.RData")){
    save(TCGA_TMB, file = "results/TCGA_TMB.RData")
}

# rm(list = grep("tcga_*", ls(), value = TRUE))
rm(list = ls()); gc()
```

