---
title: "Antigen presentation and tumor immunogenicity in cancer immunotherapy response prediction"
author: "Shixiang Wang, Xue-Song Liu"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    toc_depth: 3
    self_contained: false
---


```{r knitr_init, echo=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
               comment="#>",
               cache = TRUE, dpi = 300)
opts_knit$set(width=75)
```


This document is compiled from an Rmarkdown file which contains all code or description necessary to reproduce the analysis for the accompanying project. Each section below describes a different component of the analysis and all numbers and figures are generated directly from the underlying data on compilation.

# Dependencies

The preprocessing step is dependent on R software and some R packages, if you have not used R yet or R is not installed, please see <https://cran.r-project.org/>.

R packages:

* [UCSCXenaTools](https://github.com/ShixiangWang/UCSCXenaTools) - download data from UCSC Xena
* GEOquery - download data from NCBI GEO database
* [tidyverse](https://www.tidyverse.org/) - operate data, plot
* data.table - operate data
* survival - built in R, used to do survival analysis 
* forestplot - plot forestplot
* survminer - plot survival fit
* pROC - ROC analysis and visualization
* [TCGAmutations](https://github.com/PoisonAlien/TCGAmutations) - download TCGA mutation data
* [DT](https://cran.r-project.org/web/packages/DT/index.html) - show data table as a table in html
* [GSVA](https://github.com/rcastelo/GSVA) - GSVA algorithm implementation
* [ggstatsplot](https://github.com/IndrajeetPatil/ggstatsplot) - plot scatter with linear fit
* [corrplot](https://cran.r-project.org/web/packages/corrplot/) - plot correlation 
* knitr, rmdformats - used to compile this file
* readxl - read xlsx data

These R packages are easily searched by internet and have no strict version requirements to reproduce the following analyses.

# Data download and preprocessing

```{r child = 'preprocessing.Rmd'}
```


# TCGA Pan-cancer analyses 

```{r child = 'tcga_pancan_analyses.Rmd'}
```

# Immunotherapy datasets analyses 

```{r child = 'icb_analyses.Rmd'}
```

# Supplementary analyses

## Random selection of APM genes or IIS genes

TCGA contains about 10,000 patients with RNASeq data, it is hard to simulate GSVA serveral times with whole data (too much computation), thus we using a sampling strategy as following:

* randomly select RNAseq data from 500 patients.
* calculate GSVA score 
    * calculate once from right genes.
    * randomly select genes (10 times) and run GSVA.
* calculate spearman correlation coefficient for normal APS and normal IIS, random APS and normal IIS, normal APS and random IIS, respectively. The latter two have ten values, we use their mean.
* repeat the process 100 times.

The GSVA score calculation is implemented by script `code/random_simulation.R` (this will take much time to finish). Next we calculate correlation using following commands.

```{r}
load("results/randomRes/normal_results.RData")
load("results/randomRes/random_res.RData")
```


```{r}
source("../code/functions.R")
cor_res = data.frame(stringsAsFactors = FALSE)
for (i in 1:100){
    normal = normal_res[[i]][[1]]
    normal = calc_TisIIs(normal)
    
    r1 = as.numeric(cor.test(normal$APM, normal$IIS, method = "spearman")$estimate)
    
    r2 = c() # normal APM, random IIS
    r3 = c() # random APM, normal IIS
    for (j in 1:10){
        random = random_res[[i]][[j]][[1]]
        tryCatch(
            {
                random = calc_TisIIs(random)
                r_random1 = as.numeric(cor.test(normal$APM, random$IIS, method = "spearman")$estimate)
                r2 = c(r2, r_random1)
                r_random2 = as.numeric(cor.test(random$APM, normal$IIS, method = "spearman")$estimate)
                r3 = c(r3, r_random2)
            },
            error = function(e) {
                NA
            })
    }
    r2 = mean(r2, na.rm = TRUE)
    r3 = mean(r3, na.rm = TRUE)
    cor_res = rbind(cor_res, c(r1, r2, r3))
}

colnames(cor_res) = c("normal", "random_IIS", "random_APS")

write_tsv(cor_res, "results/randomGSVA.tsv")
```

Now, we plot spearman correlation efficient of 3 types.

```{r}
library(tidyverse)
library(cowplot)
cor_res = read_tsv("results/randomGSVA.tsv", col_types = cols())

cor_res_long = gather(cor_res, key="Type", value="SpearmanCoeff")
cor_res_long$Type = factor(cor_res_long$Type)

p = ggplot(cor_res_long, aes(x=SpearmanCoeff, color=Type))  + 
    stat_density(geom = "line") + 
    xlab("Spearman correlation coefficient") + ylab("Density") +
    guides(color=guide_legend(title = NULL)) + 
    scale_color_discrete(labels=c("r(APS, IIS)", "r(APSr, IIS)", "r(APS, IISr"))
p
```

```{r}
# save
save_plot("random_simulation.pdf", plot = p, base_aspect_ratio = 1.6)
```


We can clearly see that if we change gene list for APS or IIS calculation, the strong positive correlation between APS and IIS will no longer exist.  

## Explore correlation between ORR and TIGS from TCGA data 

A reviewer commented on our analysis which showed in section "TIGS definition and performance":

> it appears that the authors lumped together data from different data sets (e.g. reference 26 for tumor mutational burden and TCGA data for expression)

This analysis we adopted same idea from [*Tumor Mutational Burden and Response Rate to PD-1 inhibition*](https://www.ncbi.nlm.nih.gov/pubmed/29348162). To further check reliability of this analysis, here we use gene expression and TMB data from TCGA and ORR data from clinical literatures.

Firstly we load TCGA data and calculate TIGS score.

```{r}
library(dplyr)
load("results/TCGA_ALL.RData")

tcga_all = tcga_all %>% 
    mutate(nAPM = (APM - min(APM, na.rm = TRUE))/ (max(APM, na.rm = TRUE) - min(APM, na.rm = TRUE)),
           nTMB = TMB_NonsynVariants / 38, 
           TIGS = log(nTMB+1) * nAPM)
```

Next we calculate median of APS, TMB, TIGS and their corresponding sample counts for each tumor type.

```{r}
tcga_tigs = tcga_all %>% 
  dplyr::filter(!is.na(Project)) %>% 
  group_by(Project) %>% 
  summarise(Patients_TIGS=sum(!is.na(TIGS)), TIGS=median(TIGS, na.rm = TRUE),
            Patients_APS=sum(!is.na(nAPM)), APS=median(nAPM, na.rm = TRUE),
            Patients_TMB=sum(!is.na(nTMB)), TMB=median(nTMB, na.rm = TRUE))
DT::datatable(tcga_tigs)
```

We then save this data and merge them with ORR data according to corresponding tumor types.

```{r, eval=FALSE}
readr::write_csv(tcga_tigs, path = "results/TCGA_TIGS_byProject2.csv")
```

After merging data, we load data into R.

```{r}
tcga_orr = readr:::read_csv("results/TCGA_TIGS_byProject2.csv")
```

Now we plot figures.

```{r}
library(ggplot2)
require(ggrepel)
require(scales)
```

### APS

```{r}
ggplot(filter(tcga_orr, !is.na(ORR)), aes(x=APS, y=ORR)) + 
    geom_point(aes(color=Patients_APS, size=Patients_ORR)) + 
    geom_smooth(method="lm", se=T)  + 
    geom_text_repel(aes(label=Project), size=3)  +
    labs(x="Median Normalized APM Score ", y="Objective Response Rate (%)",
         size="Objective Response Rate\n(no. of patients evaluated)", 
         color = "APM Score\n(no. of tumor analyzed)") +
    scale_size_continuous(breaks = c(50, 100, 500, 1000)) +
    scale_color_gradientn(colours = RColorBrewer::brewer.pal(5, name="OrRd")[-1],
                          breaks = c(50, 200, 500, 1000)) +
    theme_bw() +
    guides(
    color = guide_colorbar(order = 1)
  )
```

Significant correlation is observed.

```{r}
lm(ORR ~ APS, filter(tcga_orr, !is.na(ORR))) %>% summary() 
```

R and R square can be given as:

```{r}
# R
filter(tcga_orr, !is.na(ORR)) %>% summarise(r = cor(ORR, APS)) %>% pull(r)
# R square
lm(ORR ~ APS, filter(tcga_orr, !is.na(ORR))) %>% summary()  %>% .$r.squared
```

### TMB

```{r}
ggplot(filter(tcga_orr, !is.na(ORR)), aes(x=TMB, y=ORR)) + 
    geom_point(aes(color=Patients_TMB, size=Patients_ORR)) + 
    geom_smooth(method="lm", se=T)  + 
    geom_text_repel(aes(label=Project), size=3)  +
    labs(x="Median No. of Coding Somatic Mutation per MB", y="Objective Response Rate (%)",
         size="Objective Response Rate\n(no. of patients evaluated)",
         color = "Tumor Mutational Burden\n(no. of tumor analyzed)") +
    scale_x_continuous(trans = log_trans(),
                       breaks = c(2, 10, 20, 30, 40, 50),
                       labels = c(2, 10, 20, 30, 40, 50)) +
    scale_size_continuous(breaks = c(50, 100, 500, 1000)) +
    scale_color_gradientn(colours = RColorBrewer::brewer.pal(5, name="OrRd")[-1],
                          breaks = c(100, 1000, 5000, 10000)) +
    theme_bw() + 
  guides(
    color = guide_colorbar(order = 1)
  )
```

Significant correlation is observed.

```{r}
lm(ORR ~ TMB, filter(tcga_orr, !is.na(ORR))) %>% summary() 
```

R and R square can be given as:

```{r}
# R
filter(tcga_orr, !is.na(ORR)) %>% summarise(r = cor(ORR, TMB)) %>% pull(r)
# R square
lm(ORR ~ TMB, filter(tcga_orr, !is.na(ORR))) %>% summary()  %>% .$r.squared
```

### TIGS


```{r}
ggplot(filter(tcga_orr, !is.na(ORR)), aes(x=TIGS, y=ORR)) + 
    geom_point(aes(size=Patients_ORR)) + 
    geom_smooth(method="lm", se=T)  + 
    geom_text_repel(aes(label=Project), size=3)  +
    labs(x="Tumor Immunogenicity Score ", y="Objective Response Rate (%)",
         size="Objective Response Rate\n(no. of patients evaluated)") +
    scale_size_continuous(breaks = c(50, 100, 500, 1000)) +
    theme_bw() 
```

Significant correlation is observed.

```{r}
lm(ORR ~ TIGS, filter(tcga_orr, !is.na(ORR))) %>% summary() 
```

R and R square can be given as:

```{r}
# R
filter(tcga_orr, !is.na(ORR)) %>% summarise(r = cor(ORR, TIGS)) %>% pull(r)
# R square
lm(ORR ~ TIGS, filter(tcga_orr, !is.na(ORR))) %>% summary()  %>% .$r.squared
```


These data are consistent with our previous results and even better, which may be caused by fewer included tumor types.


## Predict ORR for other tumor types

Using constructed linear formula, we can predict ORR of tumor types which (median) TIGS available.

```{r}
sm_data = read_csv("../data/summary_data_new_20190411.csv", col_types = cols())
sm_data$TIGS = sm_data$Pool_APM * log(sm_data$Pool_TMB + 1)

fit = lm(Pool_ORR ~ TIGS, filter(sm_data, !is.na(Pool_ORR)))
fit
```

For example , we can predict ORR for uterine corpus endometrial carcinoma (UCEC) on the basis of median TIGS of 0.7.


```{r}
nd = data.frame(TIGS = c(0.7))
predict(fit, newdata = nd, interval = "confidence")
```

